---
title: "Q1 - Project HHH Data Preprocessing"
output: html_notebook
---

## Install and load packates

```{r message=FALSE}
library(dplyr)
library(tidyr)
library(readr)
library(readxl)
library(ggplot2)
library(purrr)
library(stringr)
library(naniar)
library(janitor)
```


## Import datasets for merging

```{r}
demo <- readxl::read_xlsx("/Users/cyap/The University of Texas Health Science Center at Houston/SPH THCI Eval - General/Data Collection/Quantitative/Plano ISD/Data/2021-06-16/Evaluation Data Collection Plan-Collin-PlanoUP_Q1-FINAL.xlsx", sheet=2)

grades <- readxl::read_xlsx("/Users/cyap/The University of Texas Health Science Center at Houston/SPH THCI Eval - General/Data Collection/Quantitative/Plano ISD/Data/2021-06-16/Evaluation Data Collection Plan-Collin-PlanoUP_Q1-FINAL.xlsx", sheet=3)
```


## Rename columns and values

* empty spaces in column names will be replaced with "_"
* all values will be in lowercase
* all leading and trailing empty spaces will be removed
* double spaces will be replaced with just one empty space

```{r}
demo <- demo %>%
  mutate(
    across(
      .cols = where(is.character),
      function(x) {
        x <- str_to_lower(x) # Make lowercase
        x <- str_trim(x) # Remove leading and trailing spaces
        x <- str_replace_all(x, "[[:blank:]]{2,}", " ") # Remove double spaces
      }
    )
  )

grades <- grades %>%
  mutate(
    across(
      .cols = where(is.character),
      function(x) {
        x <- str_to_lower(x) # Make lowercase
        x <- str_trim(x) # Remove leading and trailing spaces
        x <- str_replace_all(x, "[[:blank:]]{2,}", " ") # Remove double spaces
      }
    )
  )

demo <- clean_names(demo)
grades <- clean_names(grades)
```

```{r}
demo <- demo %>% rename(student_id = research_id)
```


## View datasets

```{r}
# show duplicated subject id from the demographics tables
n_occur <- data.frame(table(demo$student_id))
n_occur[n_occur$Freq > 1,] #tells you which ids occurred more than once.
demo[demo$student_id %in% n_occur$Var1[n_occur$Freq > 1],]

n_occur <- data.frame(table(grades$student_id))
n_occur[n_occur$Freq > 1,] #tells you which ids occurred more than once.
grades[grades$student_id %in% n_occur$Var1[n_occur$Freq > 1],]

# delete columns that contain all NAs
# demo <- demo[colSums(!is.na(demo)) > 0]
# grades <- grades[colSums(!is.na(grades)) > 0]
```

```{r}
# order by ID and date
demo <- demo[order(demo$student_id, demo$enrollment_date), ] # 42 records
grades <- grades[order(grades$student_id), ] # 22 records

# delete rows if participant ID is NA
demo <- demo[!is.na(demo$student_id),]
grades <- grades[!is.na(grades$student_id),]

## rename columns
grades <- grades %>% 
  rename(
    avg_grade_dec2020 = average_grade_sm1_dec_2020,
    avg_grade_may2021 = average_grade_sm2_may_2021,
    avg_grade_dec2021 = average_grade_sm1_dec_2021,
    avg_grade_may2022 = average_grade_sm2_may_2022,
    avg_grade_dec2022 = average_grade_sm1_dec_2022,
  )
```


## Rename columns in grades dataset
```{r}
grades <- grades %>%
  pivot_longer(
    cols         = starts_with("avg_grade"),
    names_to     = "date",
    names_prefix = "avg_grade_",
    values_to    = "avg_school_grade"
) %>% 
  mutate(
    avg_school_grade_year =
        case_when(
        str_detect(date, "2020") ~ as.integer(2020),
        str_detect(date, "2021") ~ as.integer(2021),
        str_detect(date, "2022") ~ as.integer(2022)
      ),
    avg_school_grade_month =
      case_when(
        str_detect(date, "dec") ~ as.integer(12),
        str_detect(date, "may") ~ as.integer(5)
      )
  ) %>% select(-c(date))
```


## Merge grades dataset with demographics (one-to-many relationship)

```{r}
main <- demo %>% 
  left_join(grades, by = c("student_id" ="student_id")) %>% select(-c(plano_up_enrollment_date, grade, school))
```

```{r}
main <- demo
```

```{r}
main$referral_sources[main$student_id=="pu977"] <- "counselor"
main$campus[main$student_id=="pu977"] <- "bowman middle school"
```

```{r}
main$transfer <- 0
main$transfer[main$referral_date == "transfer"] <- 1
main$referral_date[main$referral_date=="transfer"] <- NA

main$referral_date <- as.numeric(as.character(main$referral_date))
main$referral_date <- as.Date(main$referral_date, origin="1899-12-30")

na_strings <- c("n/a", "na") # can add more if found
main <- main %>% naniar::replace_with_na_all(condition= ~.x %in% na_strings)
```

```{r}
main$gad_7_date_0 <- format(main$gad_7_date_0, format="%Y-%m-%d")
main$gad_7_date_3 <- format(main$gad_7_date_3, format="%Y-%m-%d")
main$enrollment_date <- format(main$enrollment_date, format="%Y-%m-%d")
main$usda_date_0  <- format(main$usda_date_0, format="%Y-%m-%d")
main$phq_a_date_0 <- format(main$phq_a_date_0, format="%Y-%m-%d")
main$phq_a_date_3 <- format(main$phq_a_date_3, format="%Y-%m-%d")
main$phq_a_date_6 <- format(main$phq_a_date_6, format="%Y-%m-%d")
main$phq_a_date_9 <- format(main$phq_a_date_9, format="%Y-%m-%d")
main$needs_assessment_date_0 <- format(main$needs_assessment_date_0, format="%Y-%m-%d")
```

```{r}
# rearrange columns
main <- main %>% select(student_id, age, sex, race, ethnicity, grade_level, 
campus, referral_sources, referral_date, transfer, everything())

demo <- main %>% select(student_id, age, sex, race, ethnicity, grade_level, 
campus, referral_sources, referral_date, transfer, plano_up_cycle_1, 
enrollment_date)
```

# for Sunil 

```{r}
measures <- main %>% select(student_id, gad_7_date_0, gad_7_score_0, 
phq_a_date_0, phq_a_score_0, usda_date_0, usda_score_0, 
gad_7_date_3, gad_7_score_3, phq_a_date_3, phq_a_score_3, 
usda_date_3, usda_score_3, gad_7_date_6, gad_7_score_6, 
phq_a_date_6, phq_a_score_6, usda_date_6, usda_score_6, 
gad_7_date_9, gad_7_score_9, phq_a_date_9, phq_a_score_9, 
usda_date_9, usda_score_9, gad_7_date_12, gad_7_score_12, 
phq_a_date_12, phq_a_score_12, usda_date_12, usda_score_12, 
needs_assessment_date_0, need_1_rental_housing_0, need_1_status_0, 
need_2_utility_assistance_0, need_2_status_0, need_3_food_assistance_0, 
need_3_status_0, need_4_clothing_0, need_4_status_0, need_5_mental_health_support_0, 
need_5_status_0, need_6_parent_education_0, need_6_status_0, 
need_7_life_skills_training_0, need_7_status_0, need_8_medical_care_0, 
need_8_status_0, need_9_dental_vision_care_0, need_9_status_0, 
program_status_0, needs_assessment_date_3, need_1_rental_housing_3, 
need_1_status_3, need_2_utility_assistance_3, need_2_status_3, 
need_3_food_assistance_3, need_3_status_3, need_4_clothing_3, 
need_4_status_3, need_5_mental_health_support_3, need_5_status_3, 
need_6_parent_education_3, need_6_status_3, need_7_life_skills_training_3, 
need_7_status_3, need_8_medical_care_3, need_8_status_3, 
need_9_dental_vision_care_3, need_9_status_3, program_status_3, 
needs_assessment_date_6, need_1_rental_housing_6, need_1_status_6, 
need_2_utility_assistance_6, need_2_status_6, need_3_food_assistance_6, 
need_3_status_6, need_4_clothing_6, need_4_status_6, need_5_mental_health_support_6, 
need_5_status_6, need_6_parent_education_6, need_6_status_6, 
need_7_life_skills_training_6, need_7_status_6, need_8_medical_care_6, 
need_8_status_6, need_9_dental_vision_care_6, need_9_status_6, 
program_status_6, needs_assessment_date_9, need_1_rental_housing_9, 
need_1_status_9, need_2_utility_assistance_9, need_2_status_9, 
need_3_food_assistance_9, need_3_status_9, need_4_clothing_9, 
need_4_status_9, need_5_mental_health_support_9, need_5_status_9, 
need_6_parent_education_9, need_6_status_9, need_7_life_skills_training_9, 
need_7_status_9, need_8_medical_care_9, need_8_status_9, 
need_9_dental_vision_care_9, need_9_status_9, program_status_9, 
needs_assessment_date_12, need_1_rental_housing_12, need_1_status_12, 
need_2_utility_assistance_12, need_2_status_12, need_3_food_assistance_12, 
need_3_status_12, need_4_clothing_12, need_4_status_12, 
need_5_mental_health_support_12, need_5_status_12, need_6_parent_education_12, 
need_6_status_12, need_7_life_skills_training_12, need_7_status_12, 
need_8_medical_care_12, need_8_status_12, need_9_dental_vision_care_12, 
need_9_status_12, program_status_12)
```

```{r}
measures_reserve <- measures
measures <- measures_reserve
```

```{r}
measures_gad7_score <- measures %>% select(student_id, contains("gad_7_score"))
measures_gad7_score_long <- measures_gad7_score %>% 
  pivot_longer(
    cols         = starts_with("gad_7_score"),
    names_to     = "timepoint",
    names_prefix = "gad_7_score_",
    values_to    = "gad7_score"
  )

rm(measures_gad7_score)

measures_gad7_date <- measures %>% select(student_id, contains("gad_7_date"))
measures_gad7_date_long <- measures_gad7_date %>% 
  pivot_longer(
    cols         = starts_with("gad_7_date_"),
    names_to     = "timepoint",
    names_prefix = "gad_7_date_",
    values_to    = "gad7_date"
  )
rm(measures_gad7_date)
measures_phqa_score <- measures %>% select(student_id, contains("phq_a_score"))
measures_phqa_score_long <- measures_phqa_score %>% 
  pivot_longer(
    cols         = starts_with("phq_a_score"),
    names_to     = "timepoint",
    names_prefix = "phq_a_score_",
    values_to    = "phqa_score"
  )
rm(measures_phqa_score)

measures_phqa_date <- measures %>% select(student_id, contains("phq_a_date"))
measures_phqa_date_long <- measures_phqa_date %>% 
  pivot_longer(
    cols         = starts_with("phq_a_date"),
    names_to     = "timepoint",
    names_prefix = "phq_a_date_",
    values_to    = "phqa_date"
  )
rm(measures_phqa_date)
measures_usda_score <- measures %>% select(student_id, contains("usda_score"))
measures_usda_score_long <- measures_usda_score %>% 
  pivot_longer(
    cols         = starts_with("usda_score"),
    names_to     = "timepoint",
    names_prefix = "usda_score_",
    values_to    = "usda_score"
  )
rm(measures_usda_score)
measures_usda_date <- measures %>% select(student_id, contains("usda_date"))
measures_usda_date_long <- measures_usda_date %>% 
  pivot_longer(
    cols         = starts_with("usda_date"),
    names_to     = "timepoint",
    names_prefix = "usda_date_",
    values_to    = "usda_date"
  )
rm(measures_usda_date)

outcomes <- measures_gad7_date_long %>% 
  left_join(measures_gad7_score_long, by = c("student_id"="student_id", "timepoint"="timepoint")) %>% 
  left_join(measures_phqa_date_long, by = c("student_id"="student_id", "timepoint"="timepoint")) %>%
  left_join(measures_phqa_score_long, by = c("student_id"="student_id", "timepoint"="timepoint")) %>%
  left_join(measures_usda_date_long, by = c("student_id"="student_id", "timepoint"="timepoint")) %>%
  left_join(measures_usda_score_long, by = c("student_id"="student_id", "timepoint"="timepoint"))

rm(measures_gad7_date_long, measures_gad7_score_long, measures_phqa_date_long, measures_phqa_score_long, measures_usda_date_long, measures_usda_score_long, n_occur)
```


```{r}
needs_ass_date <- measures %>% select(student_id, contains("needs_assessment_date"))
needs_ass_date_long <- needs_ass_date %>% 
  pivot_longer(
    cols         = starts_with("needs_assessment_date"),
    names_to     = "timepoint",
    names_prefix = "needs_assessment_date_",
    values_to    = "needs_assessment_date"
  )

needs_ass_status <- measures %>% select(student_id, contains("need_"))

need1_yn <- needs_ass_status %>% select(student_id, contains("need_1_rental_housing"))
need1_yn_long <- need1_yn %>% 
  pivot_longer(
    cols         = starts_with("need_1_rental_housing"),
    names_to     = "timepoint",
    names_prefix = "need_1_rental_housing_",
    values_to    = "need_housing"
  )

need1_status <- needs_ass_status %>% select(student_id, contains("need_1_status"))
need1_status_long <- need1_status %>% 
  pivot_longer(
    cols         = starts_with("need_1_status"),
    names_to     = "timepoint",
    names_prefix = "need_1_status_",
    values_to    = "need_housing_status"
  )

need1 <- need1_yn_long %>% 
  left_join(need1_status_long, by = c("student_id"="student_id", "timepoint"="timepoint"))



need2 <- needs_ass_status %>% select(student_id, contains("need_2"))
need2_yn <- needs_ass_status %>% select(student_id, contains("need_2_utility_assistance"))
need2_yn_long <- need2_yn %>% 
  pivot_longer(
    cols         = starts_with("need_2_utility_assistance"),
    names_to     = "timepoint",
    names_prefix = "need_2_utility_assistance_",
    values_to    = "need_utility_assistance"
  )

need2_status <- needs_ass_status %>% select(student_id, contains("need_2_status"))
need2_status_long <- need2_status %>% 
  pivot_longer(
    cols         = starts_with("need_2_status"),
    names_to     = "timepoint",
    names_prefix = "need_2_status_",
    values_to    = "need_utility_assistance_status"
  )

need2 <- need2_yn_long %>% 
  left_join(need2_status_long, by = c("student_id"="student_id", "timepoint"="timepoint"))




need3 <- needs_ass_status %>% select(student_id, contains("need_3"))
need3_yn <- needs_ass_status %>% select(student_id, contains("need_3_food_assistance"))
need3_yn_long <- need3_yn %>% 
  pivot_longer(
    cols         = starts_with("need_3_food_assistance"),
    names_to     = "timepoint",
    names_prefix = "need_3_food_assistance_",
    values_to    = "need_food_assistance"
  )

need3_status <- needs_ass_status %>% select(student_id, contains("need_3_status"))
need3_status_long <- need3_status %>% 
  pivot_longer(
    cols         = starts_with("need_3_status"),
    names_to     = "timepoint",
    names_prefix = "need_3_status_",
    values_to    = "need_food_assistance_status"
  )

need3 <- need3_yn_long %>% 
  left_join(need3_status_long, by = c("student_id"="student_id", "timepoint"="timepoint"))



need4 <- needs_ass_status %>% select(student_id, contains("need_4"))
need4_yn <- needs_ass_status %>% select(student_id, contains("need_4_clothing"))
need4_yn_long <- need4_yn %>% 
  pivot_longer(
    cols         = starts_with("need_4_clothing"),
    names_to     = "timepoint",
    names_prefix = "need_4_clothing_",
    values_to    = "need_clothing"
  )

need4_status <- needs_ass_status %>% select(student_id, contains("need_4_status"))
need4_status_long <- need4_status %>% 
  pivot_longer(
    cols         = starts_with("need_4_status"),
    names_to     = "timepoint",
    names_prefix = "need_4_status_",
    values_to    = "need_clothing_status"
  )

need4 <- need4_yn_long %>% 
  left_join(need4_status_long, by = c("student_id"="student_id", "timepoint"="timepoint"))


need5 <- needs_ass_status %>% select(student_id, contains("need_5"))
need5_yn <- needs_ass_status %>% select(student_id, contains("need_5_mental_health_support"))
need5_yn_long <- need5_yn %>% 
  pivot_longer(
    cols         = starts_with("need_5_mental_health_support"),
    names_to     = "timepoint",
    names_prefix = "need_5_mental_health_support_",
    values_to    = "need_mental_health"
  )

need5_status <- needs_ass_status %>% select(student_id, contains("need_5_status"))
need5_status_long <- need5_status %>% 
  pivot_longer(
    cols         = starts_with("need_5_status"),
    names_to     = "timepoint",
    names_prefix = "need_5_status_",
    values_to    = "need_mental_health_status"
  )

need5 <- need5_yn_long %>% 
  left_join(need5_status_long, by = c("student_id"="student_id", "timepoint"="timepoint"))


need6 <- needs_ass_status %>% select(student_id, contains("need_6"))
need6_yn <- needs_ass_status %>% select(student_id, contains("need_6_parent_education"))
need6_yn_long <- need6_yn %>% 
  pivot_longer(
    cols         = starts_with("need_6_parent_education"),
    names_to     = "timepoint",
    names_prefix = "need_6_parent_education_",
    values_to    = "need_parent_education"
  )

need6_status <- needs_ass_status %>% select(student_id, contains("need_6_status"))
need6_status_long <- need6_status %>% 
  pivot_longer(
    cols         = starts_with("need_6_status"),
    names_to     = "timepoint",
    names_prefix = "need_6_status_",
    values_to    = "need_parent_education_status"
  )

need6 <- need6_yn_long %>% 
  left_join(need6_status_long, by = c("student_id"="student_id", "timepoint"="timepoint"))


need7 <- needs_ass_status %>% select(student_id, contains("need_7"))
need7_yn <- needs_ass_status %>% select(student_id, contains("need_7_life_skills_training"))
need7_yn_long <- need7_yn %>% 
  pivot_longer(
    cols         = starts_with("need_7_life_skills_training"),
    names_to     = "timepoint",
    names_prefix = "need_7_life_skills_training_",
    values_to    = "need_life_skills"
  )

need7_status <- needs_ass_status %>% select(student_id, contains("need_7_status"))
need7_status_long <- need7_status %>% 
  pivot_longer(
    cols         = starts_with("need_7_status"),
    names_to     = "timepoint",
    names_prefix = "need_7_status_",
    values_to    = "need_life_skills_status"
  )

need7 <- need7_yn_long %>% 
  left_join(need7_status_long, by = c("student_id"="student_id", "timepoint"="timepoint"))


need8 <- needs_ass_status %>% select(student_id, contains("need_8"))
need8_yn <- needs_ass_status %>% select(student_id, contains("need_8_medical_care"))
need8_yn_long <- need8_yn %>% 
  pivot_longer(
    cols         = starts_with("need_8_medical_care"),
    names_to     = "timepoint",
    names_prefix = "need_8_medical_care_",
    values_to    = "need_medical_care"
  )

need8_status <- needs_ass_status %>% select(student_id, contains("need_8_status"))
need8_status_long <- need8_status %>% 
  pivot_longer(
    cols         = starts_with("need_8_status"),
    names_to     = "timepoint",
    names_prefix = "need_8_status_",
    values_to    = "need_medical_care_status"
  )

need8 <- need8_yn_long %>% 
  left_join(need8_status_long, by = c("student_id"="student_id", "timepoint"="timepoint"))


need9 <- needs_ass_status %>% select(student_id, contains("need_9"))
need9_yn <- needs_ass_status %>% select(student_id, contains("need_9_dental_vision_care"))
need9_yn_long <- need9_yn %>% 
  pivot_longer(
    cols         = starts_with("need_9_dental_vision_care"),
    names_to     = "timepoint",
    names_prefix = "need_9_dental_vision_care_",
    values_to    = "need_dental_vision_care"
  )

need9_status <- needs_ass_status %>% select(student_id, contains("need_9_status"))
need9_status_long <- need9_status %>% 
  pivot_longer(
    cols         = starts_with("need_9_status"),
    names_to     = "timepoint",
    names_prefix = "need_9_status_",
    values_to    = "need_dental_vision_care_status"
  )

need9 <- need9_yn_long %>% 
  left_join(need9_status_long, by = c("student_id"="student_id", "timepoint"="timepoint"))


needs_assessment <- needs_ass_date_long %>% 
  left_join(need1, by = c("student_id"="student_id", "timepoint"="timepoint")) %>%
  left_join(need2, by = c("student_id"="student_id", "timepoint"="timepoint")) %>%
  left_join(need3, by = c("student_id"="student_id", "timepoint"="timepoint")) %>% 
  left_join(need4, by = c("student_id"="student_id", "timepoint"="timepoint")) %>% 
  left_join(need5, by = c("student_id"="student_id", "timepoint"="timepoint")) %>% 
  left_join(need6, by = c("student_id"="student_id", "timepoint"="timepoint")) %>% 
  left_join(need7, by = c("student_id"="student_id", "timepoint"="timepoint")) %>% 
  left_join(need8, by = c("student_id"="student_id", "timepoint"="timepoint")) %>% 
  left_join(need9, by = c("student_id"="student_id", "timepoint"="timepoint"))
```

```{r}
program_status <- measures %>% select(student_id, contains("program_status"))

program_status_long <- program_status %>%
  pivot_longer(
    cols         = starts_with("program_status"),
    names_to     = "timepoint",
    names_prefix = "program_status_",
    values_to    = "program_status"
  )
```

## Save merged file
```{r}
# long datasets
readr::write_csv(demo, "/Users/cyap/The University of Texas Health Science Center at Houston/SPH THCI Eval - General/Data Collection/Quantitative/Plano ISD/Analysis/Steph/long datasets/q1_planoup_demo_07062021.csv")

readr::write_csv(grades, "/Users/cyap/The University of Texas Health Science Center at Houston/SPH THCI Eval - General/Data Collection/Quantitative/Plano ISD/Analysis/Steph/long datasets/q1_planoup_grades_07062021.csv")

readr::write_csv(outcomes, "/Users/cyap/The University of Texas Health Science Center at Houston/SPH THCI Eval - General/Data Collection/Quantitative/Plano ISD/Analysis/Steph/long datasets/q1_planoup_measures_07062021.csv")

readr::write_csv(needs_assessment, "/Users/cyap/The University of Texas Health Science Center at Houston/SPH THCI Eval - General/Data Collection/Quantitative/Plano ISD/Analysis/Steph/long datasets/q1_planoup_needs_assessment_07062021.csv")

readr::write_csv(program_status_long, "/Users/cyap/The University of Texas Health Science Center at Houston/SPH THCI Eval - General/Data Collection/Quantitative/Plano ISD/Analysis/Steph/long datasets/q1_planoup_program_status_07062021.csv")

# wide version
readr::write_csv(main, "/Users/chiufengyap/The University of Texas Health Science Center at Houston/SPH THCI Eval - General/Data Collection/Quantitative/Plano ISD/Analysis/Steph/wide datasets/q1_planoup_wide_07062021.csv")
```


